Predicting tree failure likelihood for utility risk mitigation via a novel convolutional neural network

### 2021-02-02 docs/manuscript/manuscript.tex

Contact between tree parts and power lines can take several forms: tree branches can grow into lines; branches can fail and fall onto lines; whole-tree failure can occur due to uprooting or trunk failure.  A study in the state of Connecticut provides some context for the amount of economic, documenting annual disruptions of \$8.3 billion between 2005 and 2015 \cite{graziano2020wider}.  That extremely high cost occurs despite extensive efforts on the part of utilities to mitigate such disruptions through active and aggressive pruning programs that, on their own cost billions of dollars annually \cite{guggenmoos2003effects}.

Pruning, despite its high cost, has been found to be effective in reducing disruptions due to so-called ``preventable'' contact incidents between trees and power lines.  For example, in Massachusetts, where tree failure was responsible for 40\% of preventable tree-caused outages, pruning was able to improve reliability by 20\% to 30\% \cite{simpson1996treecaused}, and those results were replicated in a study in Connecticut \cite{parent2019analysis}. The efficacy of pruning has also been shown in a study of two US Gulf Coast states that showed wind-induced power outage prediction models becoming less uncertain when pruning was included in the model \citet{nateghi2014power}.

Even effective pruning cannot, however, completely eliminate tree-related outages. Failure of trees away from the right-of-way can still impact the lines and cause outages \cite{guggenmoos2003effects}. The proportion of tree failures away from the wires that cause outages varies and has not been rigorously quantified. \citet{guggenmoos2011treerelated} estimated that 95\% of tree-caused outages in the Pacific Northwest region of the US, were due to tree failure, and \citet{wismer2018targeted} reported approximately 25\% of interruptions in Illinois, USA, were caused by trees that uprooted or broke in the stem.

Predicting the likelihood of failure is an inexact science, but tree risk assessment best management practices have been developed \cite{e.thomassmiley2017best,johnw.goodfellow2020best}. Risk includes assessing the likelihood of failure, the likelihood of impact, and the severity of consequences. The likelihood of failure depends on the anticipated loads on the tree and its load-bearing capacity. The likelihood of impact depends on proximity to the target (the lines, poles, and other hardware---``infrastructure''---in the case of utility tree risk assessment), the target’s occupancy rate (which is constant for utility lines) and whether the target is sheltered, for example by neighboring trees. Severity of consequences depends on the damage done to the infrastructure and, more importantly in some cases, the economic costs and disruption associated with outages---this, in turn, is partially related to the size of the tree or tree part that fails, and how much momentum it has when it impacts the infrastructure.

Individual tree risk assessment can be costly because of the time it requires. In some situations, a less time-consuming assessment may be justified to reduce costs, i.e.\ a ``Level 1'' assessment  \cite{e.thomassmiley2017best}. Studies have shown that trees with greater risk ratings were more likely to be detected from Level 1 risk assessments conducted in a moving vehicle in Rhode Island, USA \cite{rooney2005reliability} and Florida, USA \cite{koeser2016frequency}. The utility of Level 1 assessments in these areas suggests that artificial intelligence (AI) tools may be an effective way to reduce the cost of tree risk assessment.

The method described in the paper uses convolutional neural networks (CNN) to classify images of trees among three categories of failure likelihood: probable, possible and improbably.  The data used for training, testing and illustration of the method consists of 505 tree images that have been classified by the authors according to prevailing standards employed by professional utility arborists.

The remainder of the paper provides background AI and risk assessment, utility tree risk assessment and CNNs. The goal is to further demonstrate an innovative automated approach to tree risk assessment using an AI tool that can be readily deployed for use in various locations and also continually improved through subsequent training on new datasets.

@@

A relevant application for tree species identification using a convolutional neural network (CNN) was even recently demonstrated \cite{fricker2019convolutional}. Yet, AI has not been applied to the problem of tree-utility line risk assessment---one that is complicated by the very large number of tree species to be considered, seasonal variation in tree appearance and associated risk and local meteorological conditions.

@@

The big data revolution coupled with technological advancements that have made it possible to capture and store high resolution images have raised challenges that continue to be surmounted with successively high-performing
architectures. Over the past decade, some of these efforts resulted in significant breakthroughs in performance. AlexNet \cite{krizhevsky2012imageneta}, with 5 convolutional layers and 3 dense layers---one of the largest CNNs of its time, won the ILSVRC-2012\footnote{ImageNet Large Scale Visual Recognition Challenge; held annually from 2010 through 2017.} competition with a top-5 error rate of 15.3\% and served as a landmark in the Deep Learning subdomain. \citet{zeiler2014visualizing} then introduced ZFNet, besting the performance of AlexNet, and pioneered visualization techniques that were foundational for model inference and interpretability.  In the same year, GoogLeNet, a 22-layer network, was proposed \cite{szegedy2014going}, featuring the novel ``Inception module,'' which allowed for efficiency and accuracy in a very deep network. Subsequent improvements have been proposed to the original inception framework \cite{szegedy2015rethinking,szegedy2016inceptionv4}.  VGGNet \cite{simonyan2015very} also pushed the boundaries of depth with up 19 layers, achieving state-of-the-art performance at ILSVRC-2014. Finally, ResNet \cite{he2015deep} addressed the accuracy degradation problem that arises with increasing depth in a network by succesively fitting smaller sets of layers to the residual and employing skip connections. With these innovations, an unprecedented level of depth was achieved. Implementations with with 34, 50, 101 and 152 layers were demonstrated. ResNet-152 won first place in ILSVRC-2015.

@@

\section{Data and Methods}
\subsection{Image data description}
The training dataset consists of 505 images, each having an original size of $4032\times3024$ pixels.   Images were captured over a single field season in Massachusetts, USA, between May and September 2020 to limit any potential influence of changes in tree appearance due to seasonal leaf senescence on image processing.  ESRI ArcMaps was used to randomly distribute sampling sites across the state.  Risk assessments followed the “Level 1” methods outlined in the second edition of the International Society of Arboriculture’s (ISA) Tree Risk Assessment Beast Management Practices \cite{e.thomassmiley2017best} and ISA’s Utility Tree Risk Assessment Best Management Practices \cite{johnw.goodfellow2020best}.  This method is commonly used to assess trees in the United States.  This level assessment was selected for this study because: (1) individual risk assessments may be prohibitively expensive at higher orders, i.e. Level 2 or Level 3 \cite{e.thomassmiley2017best}, given the hundreds of thousands of trees utilities must manage across territory areas;  (2) utility rights-of-way (ROW) rights may not allow utility inspectors full access to trees in practical application of higher order risk assessment procedure if the trees are beyond the edge of the ROW \cite{johnw.goodfellow2020best}; (3) studies have shown reasonable efficacy of limited basic visual assessment techniques in identifying more severe tree defects \cite{rooney2005reliability,koeser2016frequency} leading to more severe likelihood of failure ratings.  The four categories of tree failure likelihood are defined as follows \cite{e.thomassmiley2017best}:
\begin{itemize}
\item \textbf{Improbable}: failure unlikely either during normal or extreme weather conditions 
\item \textbf{Possible}: failure expected under extreme weather conditions; but unlikely during normal weather conditions
\item \textbf{Probable}: failure expected under normal weather conditions within a given time frame
\item \textbf{Imminent}: failure has started or is most likely to occur in the near future, even if there is no significant wind or increased load. This is a rare occurrence for a risk assessor to encounter, and may require immediate action to protect people from harm
\end{itemize}
	In this study, only images of trees assigned to the lowest 3 likelihood of failure categories of \textit{improbable, possible, \textit{and} probable} were used due to the rarity of the \textit{imminent} category trees.

@@

To achieve robustness in training, and given the relatively small number of training images, we randomly cropped each image on either axis to 3024 $\times$ 3024 pixels, generating five instances for each one. Thus, we increased the size of our training set from 505 to 2525 images. Further, we performed horizontal flipping with a 50\% probability on each of the generated images. For efficiency, converted the images to grayscale and scaled the pixel values from 0 to 1. Finally, we downsampled the images to the following resolutions: $64 \times 64$, $128 \times 128$, $224 \times 224$ and $384\times 384$, creating a training set for each case.

>>>>

Contact between tree parts and power lines can take several forms: tree branches can grow into lines; branches can fail and fall onto lines; whole-tree failure can occur due to uprooting or trunk failure.  A study in the state of Connecticut, USA provides some context for the amount of economic disruption, documenting annual disruptions of \$8.3 billion between 2005 and 2015 \cite{graziano2020wider}.  That extremely high cost occurred despite extensive efforts on the part of utilities to mitigate conflicts between trees and power lines through active and aggressive pruning programs that, on their own cost billions of dollars annually \cite{guggenmoos2003effects}.

Pruning, despite its high cost, has been found to be effective in reducing disruptions due to so-called ``preventable'' contact incidents between trees and power lines.  For example, in Massachusetts, USA, where tree failure was responsible for 40\% of preventable tree-caused outages, pruning was able to improve reliability by 20\% to 30\% \cite{simpson1996treecaused}, and those results were replicated in a study in Connecticut \cite{parent2019analysis}. The efficacy of pruning has also been shown in a study of two states in the Gulf Coast region of the USA that showed wind-induced power outage prediction models becoming less uncertain when pruning was included in the model \citet{nateghi2014power}.

Even effective pruning cannot, however, completely eliminate tree-caused outages. Failure of trees away from the right-of-way can still impact the lines and cause outages \cite{guggenmoos2003effects}. The proportion of tree failures away from the wires that causes outages varies and has not been rigorously quantified. \citet{guggenmoos2011treerelated} estimated that 95\% of tree-caused outages in the Pacific Northwest region of the USA, were due to tree failure, and \citet{wismer2018targeted} reported approximately 25\% of interruptions in Illinois, USA, were caused by trees that uprooted or broke in the stem.

Predicting the likelihood of failure is an inexact science, but tree risk assessment best management practices have been developed \cite{e.thomassmiley2017best,johnw.goodfellow2020best}. Risk includes assessing the likelihood of failure, the likelihood of impact, and the severity of consequences. The likelihood of failure depends on the anticipated loads on the tree and its load-bearing capacity. The likelihood of impact depends on proximity to the target (the lines, poles, and other hardware---``infrastructure''---in the case of utility tree risk assessment), the target’s occupancy rate (which is constant for utility lines) and whether the target is sheltered, for example by neighboring trees. Severity of consequences depends on the damage done to the infrastructure---which, in turn, is partially related to the size of the tree or tree part that fails, and how much momentum it has when it impacts the infrastructure---and, more importantly in some cases, the economic costs and disruption associated with outages.

Individual tree risk assessment can be costly because of the time it requires. In some situations, a less time-consuming assessment may be justified to reduce costs, i.e.\ a ``Level 1'' assessment  \cite{e.thomassmiley2017best}. Studies have shown that trees with greater risk ratings were more likely to be detected from Level 1 risk assessments conducted in a moving vehicle in Rhode Island, USA \cite{rooney2005reliability} and Florida, USA \cite{koeser2016frequency}. The utility of Level 1 assessments in these states suggests that artificial intelligence (AI) tools may be an effective way to reduce the cost of tree risk assessment.

The method described in the paper uses convolutional neural networks (CNN) to classify images of trees among three categories of failure likelihood: probable, possible, and improbable.  The data used for training, testing and illustration of the method consists of 505 tree images that have been classified by the authors according to prevailing standards employed by professional utility arborists.

The remainder of the paper provides a brief history and background of AI and its use in earthquake risk assessment and tree identification (section 2); describes the methods used to train and validate a novel CNN to categorize likelihood of tree failure (section 3); and presents and discusses the output of the novel CNN (sections 4 and 5). The goal is to further demonstrate an innovative automated approach to tree risk assessment using an AI tool that can be readily deployed for use in various locations and also continually improved through subsequent training on new datasets.

@@

A relevant application for tree species identification using a convolutional neural network (CNN) was recently demonstrated \cite{fricker2019convolutional}. Yet, AI has not been applied to the problem of tree-utility line risk assessment---one that is complicated by the very large number of tree species to be considered, seasonal variation in tree appearance and associated risk and local meteorological conditions.

@@

The big data revolution coupled with technological advancements that have made it possible to capture and store high resolution images have raised challenges that continue to be surmounted with successively high-performing
architectures. Over the past decade, some of these efforts resulted in significant breakthroughs in performance. AlexNet \cite{krizhevsky2012imageneta}, with 5 convolutional layers and 3 dense layers---one of the largest CNNs of its time, won the ILSVRC-2012\footnote{ImageNet Large Scale Visual Recognition Challenge; held annually from 2010 through 2017.} competition with a top-5 error rate of 15.3\% and served as a landmark in the Deep Learning subdomain. \citet{zeiler2014visualizing} then introduced ZFNet, besting the performance of AlexNet, and pioneered visualization techniques that were foundational for model inference and interpretability.  In the same year, GoogLeNet, a 22-layer network, was proposed \cite{szegedy2014going}, featuring the novel ``Inception module,'' which allowed for efficiency and accuracy in a very deep network. Subsequent improvements have been proposed to the original inception framework \cite{szegedy2015rethinking,szegedy2016inceptionv4}.  VGGNet \cite{simonyan2015very} also pushed the boundaries of depth with up 19 layers, achieving state-of-the-art performance at ILSVRC-2014. Finally, ResNet \cite{he2015deep} addressed the accuracy degradation problem that arises with increasing depth in a network by successively fitting smaller sets of layers to the residual and employing skip connections. With these innovations, an unprecedented level of depth was achieved. Implementations with with 34, 50, 101 and 152 layers were demonstrated. ResNet-152 won first place in ILSVRC-2015.

@@

\section{Data and Methods}
\subsection{Image data description}
The training dataset consisted of 505 images, each having an original size of $4032\times3024$ pixels.   Images were captured over a single field season in Massachusetts, USA, between May and September 2020 to limit any potential influence of changes in tree appearance due to seasonal leaf senescence on image processing.  ESRI ArcMaps was used to randomly distribute sampling sites across the state.  Risk assessments followed the “Level 1” methods outlined in the second edition of the International Society of Arboriculture’s (ISA) Tree Risk Assessment Best Management Practices \cite{e.thomassmiley2017best} and ISA’s Utility Tree Risk Assessment Best Management Practices \cite{johnw.goodfellow2020best}.  This method is commonly used to assess trees in the United States.  A level 1 assessment was selected for this study because: (1) individual risk assessments may be prohibitively expensive at higher orders, i.e. Level 2 or Level 3 \cite{e.thomassmiley2017best}, given the hundreds of thousands of trees utilities must manage across territory areas;  (2) utility rights-of-way (ROW) easements may not allow utility inspectors full access to trees in practical application of higher order risk assessment procedure if the trees are beyond the edge of the ROW \cite{johnw.goodfellow2020best}; (3) studies have shown reasonable efficacy of limited basic visual assessment techniques in identifying more severe tree defects \cite{rooney2005reliability,koeser2016frequency} leading to greater likelihood of failure ratings.  The four categories of likelihood of tree failure, which are always considered in a stated time frame, are defined as follows \cite{e.thomassmiley2017best}:
\begin{itemize}
\item \textbf{Improbable}: failure unlikely either during normal or extreme weather conditions 
\item \textbf{Possible}: failure expected under extreme weather conditions; but unlikely during normal weather conditions
\item \textbf{Probable}: failure expected under normal weather conditions within a given time frame
\item \textbf{Imminent}: failure has started or is most likely to occur in the near future, even if there is no significant wind or increased load. This is a rare occurrence for a risk assessor to encounter, and may require immediate action to protect people from harm
\end{itemize}
	In this study, only images of trees assigned to the lowest 3 likelihood of failure categories of \textit{improbable, possible, \textit{and} probable} were used due to the rarity of trees in the \textit{imminent} category.

@@

To achieve robustness in training, and given the relatively small number of training images, we randomly cropped each image on either axis to 3024 $\times$ 3024 pixels, generating five instances for each one. Thus, we increased the size of our training set from 505 to 2525 images. Further, we performed horizontal flipping with a 50\% probability on each of the generated images. For efficiency, we converted the images to grayscale and scaled the pixel values from 0 to 1. Finally, we downsampled the images to the following resolutions: $64 \times 64$, $128 \times 128$, $224 \times 224$ and $384\times 384$, creating a training set for each case.
